Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-15 main.Rmd
till visuyalization minor

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ such that $C_i$ maps index set to $n_i$ levels or categories and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}, let $S_{ij}$ be a subset of the index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the choice of the granularities does.

@@

However, we need to consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ with $n$ categories and $C_2$ with $m$ categories. Each element of $C_1$ occurs approximately $T/n$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(mn)$. If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge mnk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

If there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarter-of-year. This plot does not work because every quarter does not correspond to all months of the year for example, the first quarter would not correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weeekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

Even with harmonies, visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (\ref{sec:summarize-measured}) or unevenly distributed events.

>>>>

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities $C_i$ and $v$ denotes the measurement variable. The total number of cyclic granularities would be based on the number of linear granularities considered in the hierarchy table and presence of any aperiodic cyclic granularities. Suppose we have $n$ linear granularities in the hierarchy table, $n(n-1)/2$ circular or quasi-circular cyclic granularities could be constructed. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, this big problem is broken down by focusing on two cyclic granularities at a time. Hence data sets of the form <$C_i$, $C_j$, $v$> forms the basis for exploration and analysis of the measured variable.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_l$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. Here, $A_i$'s or $B_j$'s are the levels/categories corresponding to $C_1$ and $C_2$ respectively. Let $S_{ij}$ be a subset of the index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, l}$ and $j \in {1, 2, \dots, m}$  which will lead to $lm$ data subsets. Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $lm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ be day-of-month. Then the days denoting weekends in a month would not correspond to any Business days. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularities does.

@@

However, we need to consider the effect of number of observations on the summarization even for harmonies. Suppose we have $T$ observations, and two cyclic granularities $C_1$ and $C_2$ with $l$ and $m$ categories respectively. Each element of $C_1$ occurs approximately $T/l$ times while each element of $C_2$ occurs approximately $T/m$ times. There are no empty combinations, and each combination will occur on average an equal number of times as $T\rightarrow\infty$, so the average number of observations per combination is $T/(lm)$. If we require at least $k$ observations per combination to create a meaningful summarization, then only $T \ge lmk$ will be acceptable. The value of $k$ will depend on the statistical transformation we are producing. For computing deciles, $k=10$ may be acceptable, but for density estimates, $k \ge 30$ is suggested. Rarely occurring categories such as the 366th day of the year, or the 31st day of the month can suffer from such problem.

@@

If there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes. For illustration, \autoref{fig:allFig} (a) shows the distribution of half-hourly electricity consumption through letter value plot across months of the year faceted by quarter-of-year. This plot does not work because every quarter does not correspond to all months of the year, for example, the first quarter would not correspond to the month December of an year.

@@

```{r allFig, out.width="100%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Distribution of energy consumption displayed through letter value plots. Plot (a) displays consumption across month-of-year faceted by quarter-of-year, (b) across weekday/weekend faceted by quarter-of-year and (c) across quarter-of-year faceted by weekday/weekend. Plots (b) and (c) both show harmonies since each quarter consists of both weekdays and weekends. Conversely, weekend and weekday can occur in every quarter. Plot (b) helps to compare weekends and weekdays for each quarter. It can be seen that for every quarter weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently, whereas, Plot (c) helps to compare quarters within weekdays and weekends. For example, the quartile spread of consumption shrinks/lowers from first to fourth quarter for weekdays, whereas this pattern is not true for weekends. Plot (a) shows a clash since all quarters do not correspond to all months of the year. ", fig.show = 'hold'}

@@

Visualizing distributions can be misleading if statistical transformations are performed on rarely occurring categories (\ref{sec:summarize-measured}) or unevenly distributed events.
